How to Not write Virtual Machines for Dynamic Languages

### 2007-05-27 talk/dyla2007/dyla.tex
attempts to make the OO VM (dis)advantages more readable by itemizing them. more references

\emph{Better interoperability than the C level:} Since the VM offers a standard
object model and all the languages implemented on top of it are using it, it is
easier to integrate the languages that are running on top of the VM. This
allows reuse of libraries between all the implemented languages. This is
typically the most important reason for wanting an implementation on the VM in
the first place.

\emph{Cross-platform portability:} Only the underlying VM has to be ported to
various hardware architectures and operating systems. The languages implemented
on top can then be run without change in various environments.

\emph{Better tools:} Better IDEs, debuggers and profilers.

\emph{Better implementation of low-level issues like garbage collection,
threading:} Since an OO VM is expected to be widely used and usually backed by
a company, it becomes worthwhile and possible to spend a lot of effort tuning
its garbage collector, threading model, exception support and other low-level
implementation details.

\emph{Better performance:} Similarly, object-oriented VMs usually come with a
highly tuned just-in-time compiler to make them perform well without requiring
ahead-of-time compilation to machine language. This in addition with the
previous point leads to much better performance of the languages running on top
of the VM.

\emph{Ease of implementation:} The implementation of a language on top of an OO
VM is easier because it starts at a higher level than C. Usually a
high-level language like Java or C\# is used for the language implementation,
which both offer the language implementer a much higher level of abstraction
than when implementing in C.

\emph{A single unified implementation base:} The higher level of abstraction
mentioned in the last point also helps to reduce the number of implementations
of the language. Since the implementation on top of an OO VM is more abstract
it is also easier to modify and to extend. XXX somewhat vague point

At a closer look, some of these advantages are only partially true in practice.

\emph{Better performance:} So far it seems that performance of highly dynamic
languages is not actually significantly improved on OO VMs. Jython
\footnote{Python on the Java VM} is around 5
times slower than CPython, for IronPython\footnote{Python on .NET, which
gives up on some features to improve performance}
the figures vary but it is mostly
within the same order of magnitude as CPython. The most important reason for
this is that the VM's JIT compilers are optimized for specific usage patterns
that are common in the main language of the OO VM. To get good speeds the
language implementers would have to carefully produce code that matches these
usage patterns, which is not a simple task.

\emph{Better GCs:} While this is obvious in theory, OO VMs tend to have a much
higher memory overhead to start with (XXX ref)

\emph{Cross-platform portability:} While this is true to some extend, the
situation with regard to portability is not significantly improved compared to
e.g.  C/Posix, which is relatively portable too. Also portability sometimes
comes at the price of performance, because even if the OO VM is running on a
particular hardware architecture it is not clear that the JIT is tuned for this
architecture too or working at all, which leads to significantly less
performance.

\emph{Ease of implementation:} This point is disputable. On the one hand, OO
VMs typically allow the language implementor to start at a higher level. On the
other hand they also enforce a specific object and execution model. This means
that the concepts of the implemented language need to be mapped to the
execution model of the underlying VM, which may be easy or not, depending very
much on the language in question. XXX give example where it works well.

An example where this mapping does not work too well is Prolog. While there
exist several implementations of Prolog on top of the JVM and also one on .NET,
they are not particular efficient, especially when compared to good Prolog VMs
in written in C. This is mostly because the Prolog execution model, which
involves backtracking and deep recursion does not fit the JVM and .NET very
well. Therefore the Prolog implementations on top of OO VMs resort to models
that is quite unnatural both for the OO VM and for Prolog.

On the other hand some of the benefits are real and very useful, the most
prominent being the easy interaction with the rest of the VM. Furthermore there
is better tool support and better GCs. XXX expand

@@

We implemented this idea in the PyPy project.  The dynamic language for
which we wrote an interpreter is Python.  It is a language which,
because of its size and rather intricate semantics, is a good target for
our approach, in the following sense: its previous reimplementation
(Jython for the JVM and IronPython for .NET) have each proved to be very
time-consuming to maintain.  Our implementation is by construction
easier to maintain, and extremely portable (including to C/Posix, to the
JVM and to .NET).

@@

A detailed description of this translation process is beyond the scope of the
present paper; it can be found in \cite{pypyvmconstruction}.  The actual Python
interpreter of PyPy and the results we achieved by translating it to C, LLVM
\cite{LLVM} and .NET are described in [XXX].  These results show that the
approach is practical and gives results whose performance is within the same
order of magnitude (within a factor of 2 and improving) of the hand-written,
well-tuned CPython, the C reference implementation.  These figures do not
include the spectacular speed-ups obtained in some cases by the JIT compiler
described in section \ref{subsect:dynamic_compilers}.

>>>>

\begin{itemize}
\item
\emph{Better interoperability than the C level:} Since the VM offers a standard
object model and all the languages implemented on top of it are using it, it is
easier to integrate the languages that are running on top of the VM. This
allows reuse of libraries between all the implemented languages. This is
typically the most important reason for wanting an implementation on the VM in
the first place.

\item
\emph{Cross-platform portability:} Only the underlying VM has to be ported to
various hardware architectures and operating systems. The languages implemented
on top can then be run without change in various environments.

\item
\emph{Better tools:} Better IDEs, debuggers and profilers.

\item
\emph{Better implementation of low-level issues like garbage collection,
threading:} Since an OO VM is expected to be widely used and usually backed by
a company, it becomes worthwhile and possible to spend a lot of effort tuning
its garbage collector, threading model, exception support and other low-level
implementation details.

\item
\emph{Better performance:} Similarly, object-oriented VMs usually come with a
highly tuned just-in-time compiler to make them perform well without requiring
ahead-of-time compilation to machine language. This in addition with the
previous point leads to much better performance of the languages running on top
of the VM.

\item
\emph{Ease of implementation:} The implementation of a language on top of an OO
VM is easier because it starts at a higher level than C. Usually a
high-level language like Java or C\# is used for the language implementation,
which both offer the language implementer a much higher level of abstraction
than when implementing in C.

\item
\emph{A single unified implementation base:} The higher level of abstraction
mentioned in the last point also helps to reduce the number of implementations
of the language. Since the implementation on top of an OO VM is more abstract
it is also easier to modify and to extend. XXX somewhat vague point
\end{itemize}

At a closer look, some of these advantages are only partially true in practice:

\begin{itemize}
\item
\emph{Better performance:} So far it seems that performance of highly dynamic
languages is not actually significantly improved on OO VMs. 
Jython\footnote{Python on the Java VM} is around 5
times slower than CPython, for IronPython\footnote{Python on .NET, which
gives up on some features to improve performance}
the figures vary but it is mostly
within the same order of magnitude as CPython. The most important reason for
this is that the VM's JIT compilers are optimized for specific usage patterns
that are common in the main language of the OO VM. To get good speeds the
language implementers would have to carefully produce code that matches these
usage patterns, which is not a simple task.

\item
\emph{Better GCs:} While this is obvious in theory, OO VMs tend to have a much
higher memory overhead to start with (XXX ref)

\item
\emph{Cross-platform portability:} While this is true to some extend, the
situation with regard to portability is not significantly improved compared to
e.g.  C/Posix, which is relatively portable too. Also portability sometimes
comes at the price of performance, because even if the OO VM is running on a
particular hardware architecture it is not clear that the JIT is tuned for this
architecture too or working at all, which leads to significantly less
speed.

\item
\emph{Ease of implementation:} This point is disputable. On the one hand, OO
VMs typically allow the language implementor to start at a higher level. On the
other hand they also enforce a specific object and execution model. This means
that the concepts of the implemented language need to be mapped to the
execution model of the underlying VM, which may be easy or not, depending very
much on the language in question. XXX give example where it works well.

An example where this mapping does not work too well is Prolog. While there
exist several implementations of Prolog on top of the JVM \cite{prologcafe}
\cite{DBLP:conf/jelia/Calejo04} and also one on .NET \cite{psharp},
they are not particular efficient, especially when compared to good Prolog VMs
in written in C. This is mostly because the Prolog execution model, which
involves backtracking and deep recursion does not fit the JVM and .NET very
well. Therefore the Prolog implementations on top of OO VMs resort to models
that is quite unnatural both for the OO VM and for Prolog.
\end{itemize}

On the other hand some of the benefits are real and very useful, the most
prominent being the easy interaction with the rest of the VM. Furthermore there
is better tool support and better GCs. Also for languages where the execution
model fits the OO VM well, many of the disadvantages disappear.

@@

We implemented this idea in the PyPy project \cite{pypy}.  The dynamic language
for which we wrote an interpreter is Python.  It is a language which,
because of its size and rather intricate semantics, is a good target for
our approach, in the following sense: its previous reimplementations
(Jython for the JVM and IronPython for .NET) have each proved to be very
time-consuming to maintain.  Our implementation is by construction
easier to maintain, and extremely portable (including to C/Posix, to the
JVM and to .NET).

@@

A detailed description of this translation process is beyond the scope of the
present paper; it can be found in \cite{pypyvmconstruction}.  The actual Python
interpreter of PyPy and the results we achieved by translating it to C, LLVM
\cite{LLVM} and .NET are described in \cite{architecture} \cite{translationdoc}
These results show that the
approach is practical and gives results whose performance is within the same
order of magnitude (within a factor of 2 and improving) of the hand-written,
well-tuned CPython, the C reference implementation.  These figures do not
include the spectacular speed-ups obtained in some cases by the JIT compiler
described in section \ref{subsect:dynamic_compilers}.
